---
title: "Forest Fires in Portugal - What Are The Causes?"
subtitle: "Practical Assignment of Data Mining I"
author: "By Robson Teixeira, Eduardo Rodrigues and Claudio Rocha"
date: "M:CC -- FCUP, 10/01/2021"
output:
  html_document:
    #df_print: paged
    df_print: tibble
    toc: yes
    code_folding: "hide"
    fig_caption: yes
    includes:
      in_header: header.html
    number_sections: yes
    theme: cerulean
    toc_float: yes
  html_notebook: 
    code_folding: hide
    df_print: default
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    #citation_package: natbib
    df_print: tibble
    fig_caption: yes
    fig_crop: no
    fig_height: 5
    fig_width: 7
    number_sections: yes
    toc: yes
geometry:
- top=25mm
- bottom=25mm
- left=25mm
- right=20mm
- heightrounded
header-includes:
  \usepackage{caption}
  \usepackage{fancyhdr}
  \usepackage{lmodern}
  \usepackage[detect-all]{siunitx}
highlight-style: pygments
linkcolor: blue
mainfont: Source Variable Pro
fontsize: 12pt
sansfont: Source Sans Pro
documentclass: report
urlcolor: blue
references:
- type: article-journal   
  id: geosciences10020053
  title: Mapping the Causes of Forest Fires in Portugal by Clustering Analysis
  author:
    - family: Meira Castro
      given: Ana C.
    - family: Nunes
      given: Adélia 
    - family: Sousa
      given: António
    - family: Lourenço
      given: Luciano
  publisher: Geosciences
  issued:
    year: 2020
---

```{r setup, include= F}
library(dplyr)
library(funModeling)
library(stringr)
library(chron)

# turns off scientific notation; to turn it back on, use scipen= 0
options(scipen= 999)


fires.train.path = file("data/fires2015_train.csv")
fires.test.path = file("data/fires2015_test.csv")

fires.raw <- read.csv(fires.train.path, stringsAsFactors = FALSE, 
                      na.strings = c("-", "","NA"), 
                      encoding = "UTF-8")

fires.test.raw <- read.csv(fires.test.path, stringsAsFactors = FALSE, 
                           na.strings = c("-", "","NA"), 
                           encoding = "UTF-8")

```

# Abstract {-}

# Introduction

# Problem Definition

Forest fires are a very important issue that negatively affects climate change. Typically, the causes of forest fires are those oversights, accidents and negligence committed by individuals, intentional acts and natural causes. The latter is the root cause for only a minority of the fires.

Their harmful impacts and effects on ecosystems can be major ones. Among them, we can mention the disappearance of native species, the increase in levels of carbon dioxide in the atmosphere, earth’s nutrients destroyed by the ashes, and the massive loss of wildlife.

Data mining techniques can help in the prediction of the cause of the fire and, thus, better support the decision of taking preventive measures in order to avoid tragedy. In effect, this can play a major role in resource allocation, mitigation and recovery efforts.

## Forest Fire Dataset

The ICFN - Nature and Forest Conservation Institute has the record of the list of forest fires occurred in Portugal for several years. For each fire, there is information such as the site, the alert date/hour, the extinction date/hour, the affected area and the cause type. A classifications for causes types are presented in table \@ref(tab:cause_type).

Table 1: (\#tab:cause_type) Classifications of causes of forest fires.

Cause       |Description                                           
------------|---------------------------------------------------------------------------------------------
Unknown     | absence of suficient objective evidence to determine the cause of the ignition of fire
Natural     | lightning generated in thunderstorms
Negligence  | the misguided use of fire in activities such as burning trash, mass burning of agricultural and forest fuels, fun and leisure activities; failure to properly extinguish cigarettes by smokers; the dispersal and transport of incandescent particles from chimneys; etc.
Intentional | incendiarism and arson, mostly resulting from behaviors and attitudes reacting to theconstraints of agroforestry management systems and to conflicts related to land use
Rekindling  | reburning of an area over which a fire has previously passed, but where fuel has been left that is later ignited by latent heat, sparks, or embers


The dataset used in this study was provided by ICFN, and it contains the data on reported forest fires during 2015 and its respective causes. The data are distributed in files:

* **fires2015train.csv** --- the file contain the data of 7511 reported forest fires during 2015

```{r echo= F}
glimpse(fires.raw)
```

```{r echo= F}
head(fires.raw)
```

```{r}
df_status(fires.raw)
```


Table \@ref(tab:variables) describes all variables contained in the `fires2015train.csv` of the data set.
Clearly, the type of some of variables is incorrect and inconvenient for analysis and that was taken care of in section \@ref(data-cleaning).

Table: (\#tab:variables) List of variables present in the three original files from the Flixster data set.

Variable    |Type        |Description                   |Tibble
------------|------------|------------------------------|------
age         |character   |Age of user                   |profiles
date        |character   |Date in which user rated movie|ratings
gender      |character   |User gender                   |profiles
lastlogin   |integer     |Last login numerical tag      |profiles
location    |integer     |Location identifier of user   |profiles
memberfor   |character   |Member since date             |profiles
movieid     |integer     |Movie unique identifier       |movies
movieid     |integer     |Movie unique identifier       |ratings
moviename   |character   |movie name                    |movies
profileview |character   |Numerical tag of the user     |profiles
rating      |double      |Rating of movie by user       |ratings
userid      |integer     |User unique identifier        |profiles
userid      |integer     |User unique identifier        |ratings

# Data Preparation

## Data Cleaning: 
Identifying and correcting mistakes or errors in the data.

```{r}

```

## Data Transforms
Changing the scale or distribution of variables.

```{r}

```

## Feature Engineering
Deriving new variables from available data.

```{r}

```

## Feature Selection
Identifying those input variables that are most relevant to the task.
```{r}

```


## Dimensionality Reduction
Creating compact projections of the data.



# Exploratory Data Analysis

# Conclusions, Shortcomings and Future Work

# Appendix